### 2017-06-15 manuscript/manuscript.Rmd
Draft 1 of # Introduction

abstract: >
  Recent calls for improving the reproducibility of research products in behavioral sciences have called for increased attention to the way in which researchers curate, collaborate, and share their research assets, but have failed to provide practical instructions on how to do so. In this tutorial paper, we explain how version control systems, such as the popular Git program, are especially suitable for the challenges to reproducibility faced by behavioral science regarding research asset curating, sharing, and collaborating. We then present a complete tutorial on how to use Git from the computer's command line, and with the popular graphical interface contained in the R Studio development environment. Git is easy to learn, presents an elegant solution to problems related to curating research assets, facilitates multi-site collaboration and productivity (by allowing multiple collaborators to work on the same source simultaneously) and can be implemented on common workflows with minimal extra effort. Git may also offer a suitable solution to transparent data (and material) sharing through popular online services, such as GitHub and Open Science Framework.

keywords          : "reproducibility; open science; version control; git; research methods"
wordcount         : "X"
bibliography      : ["references.bib"]
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
lang              : "english"
class             : "doc, 12pt, draftfirst"  # man / jou / doc
output            : papaja::apa6_pdf
---

@@

The lack of reproducibility is an increasingly recognized problem across scientific disciplines, and calls for changing the scientific workflow to enhance it have been published in a wide range of research areas, including Biology [@markowetz_five_2015], Ecology [@ihle_striving_2017], Neuroscience [@eglen_toward_2017] and Psychology [@munafo_manifesto_2017]. However, although exhortations to focus on reproducibility in the scientific practice are now commonplace on the pages of leading scientific journals [@baker_1500_2016], only a small minority of researchers are familiar with the tools and practices that enable implementing reproducibility in the scientific workflow. Therefore, although there now is a broad consensus that efforts to improve reproducibility are important, and even on some of the tools that may allow to do so, materials instructing researchers in using them are lacking. In this tutorial paper, we present a detailed walk-through of the git version control system for behavioral scientists.

@@

Consider the following [@ihle_striving_2017, p.2]: Have you ever found a mistake in your results without knowing what caused it? Forgot what analyses you have already done and how (and possibly, why)? Lost datasets or information about the cases or variables in a dataset? Struggled in redoing analyses when new data became available? Had difficulty understanding what data to use or how in a project that you inherited from another researcher? Answering any of these questions in the affirmative suggests that your work might benefit from improving reproducibility [@ihle_striving_2017].

But what exactly is reproducibility and why does it matter? Reproducibility can be defined as follows: "A research project is computationally reproducible if a second investigator (including you in the future) can recreate the final reported results of the project, including key quantitative findings, tables, and figures, given only a set of files and written instructions." [@kitzes_practice_2017] In the context of experimental Psychology, for example, a project would mean an experiment or a set of experiments investigating a theory or hypothesis, reported results would be a conference presentation or a submitted manuscript. In this field, key quantitative findings are usually probability values from statistical models, such as *p*-values, or tables of descriptive statistics such as (differences in) means.

Given this broad definition, here is an example of a non-reproducible project: A published journal article advertises a specific relationship between two variables, to some specified degree of uncertainty, but doesn't provide the raw data or code used to analyse it. An example of a reproducible project, on the other hand, provides a well organized package of i) the raw data supporting the claims made in the article, ii) the computer code (or steps of analysis) required to compute the summary and test statistics from the data, and iii) instructions on how to apply ii) to i), if it is not self-evident. Clearly, many projects fall in between, and can be partly reproducible--e.g. provide raw data but no code.

This definition makes clear an important distinction: reproducibility is not the same as replicability. Traditional methods and results sections in journal articles have focused on ensuring replicability (but not reproducibility) by giving detailed instructions on how to repeat the experiment and collect a data set *like* the original one. Viewed in this light, replicability is a broad methodological issue concerning the epistemology of the scientific claims; whereas reproducibility--the topic of this tutorial--concerns the minimal steps required to allow checking for the validity of the computations required to assert the scientific claim in the first place.

Although they are often conlated [TODO] reproducibility and replicability are, in fact, independent of one another. One might think that reproducibility [TODO Peng 2011?] is a requirement of replicability, but an experiment can be replicated even if the materials of the original study are not available for replicability. Likewise, the materials of a study might allow its results to be reproduced, yet the underlying scientific idea might not be replicable. Therefore, it is important to keep these two concepts distinct; in this manuscript we are interested in reproducibility.

@@

<!-- TODO work on motivating the use of Git: why is keeping track of materials an issue? Is it an issue? How does VCS help? -->

Although reproducibility might at first appear easy, it has in large parts failed in practice. For example, one survey of leading Psychology journals found that 34-58% of articles published between 1985 and 2013 had at least one inconsistently reported *p*-value [@nuijten_prevalence_2015]. Upon recalculation, the authors couldn't obtain the same *p*-values from the reported test statistics, and therefore these 34-58% of articles were, to some degree, non-reproducible.

Why, then, do so many research products fall short of the basic scientific standard of reproducibility? We suggest one reason is the poor level of organization and curation of all the research assets (data and code, usually) that play a part in creating the research product. Researchers usually have no formal methods or accepted gold standards for curating their research assets, and collaborative workflows, especially, are difficult to manage in a reproducible manner. When the complexity and amount of materials related to a project increases, it might be difficult to link analysis files to the correct data files, and track the evolution of both code and data throughout the research cycle.

In fact, the effort to organize and curate materials has been cited as an important reason for why data is so rarely shared in Psychology. In one study, researchers asked authors of 141 Psychology articles (with a total of 249 studies) to share their data for reanalysis [@wicherts_poor_2006]. 73% of the authors refused to share their data, and Wicherts et al. suggested that they did so because it takes considerable effort to clean, document and organize datasets. How could we make organizing and curating materials and data easier, so as to improve the reproducibility of our science?

## Aims of the article

Fortunately for the empirical sciences, challenges related to organizing and curating materials across time, space, and personnel have been solved to a high standard in computer science with Version Control Systems (VCS). In the remainder of this article, we introduce a popular VCS called Git, and illustrate its use in the scientific workflow with a hypothetical example project. In the tutorial below, we provide show how to use the Git VCS to curate your research materials by using text commands from the computers command line, and with a Graphical User Interface called R Studio.

@@

It is important to understand that unlike an operating system's (OS) default file viewer, such as Finder (Apple computers) or File Explorer (?? Windows), Git is not a standalone program for navigating files and folders on a computer. Instead, it adds functionality to the OSs existing file system, by making available a specific set of functions--either executed from the command line, or through a graphical user interface (GUI) / integrated development environment (IDE)--that allow taking snapshots of the project and its files, and distributing the work across multiple computers and users. To begin using Git, users must first download and install the Git software on their computers. Git is free and open source, works on Windows, Macintosh, and Linux OSs (among others), and is used by major software developers such as Microsoft, Google, and Facebook. It can be freely downloaded at <https://git-scm.com/>. Before we explain the basics of using Git, we present brief installation instructions for Windows and Apple users.

@@

[^git-version]: As of the writing of this article, the current version of Git is 2.13.1.

>>>>

abstract: >
  Recent calls for improving the reproducibility of behavioral sciences have increased attention to the ways in which researchers curate, share and collaborate on their research assets. However, these discussions have largely failed to provide practical instructions on how to do so. In this tutorial paper, we explain why version control systems, such as the popular Git program, are especially suitable for these challenges to reproducibility. We then present a tutorial on how to use Git from the computer's command line, and with the popular graphical interface contained in the R Studio development environment. This tutorial is especially written for behavioral scientists with no previous experience with version control systems. Git is easy to learn, presents an elegant solution to challenges to reproducibility related to curating research assets, facilitates multi-site collaboration and productivity (by allowing multiple collaborators to work on the same source simultaneously) and can be implemented on common behavioral science workflows with little extra effort. Git may also offer a suitable solution to transparent data (and material) sharing through popular online services, such as GitHub and Open Science Framework.

keywords          : "reproducibility; version control; git; research methods; open science"
wordcount         : "X"
bibliography      : ["references.bib"]
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
lang              : "english"
class             : "doc, 12pt, draftfirst"  # man / jou / doc
output            : papaja::apa6_pdf
---

@@

The lack of reproducibility is increasingly recognized as a problem across scientific disciplines, and calls for changing the scientific workflow to enhance reproducibility have been published in a wide range of research areas, including Biology [@markowetz_five_2015], Ecology [@ihle_striving_2017], Neuroscience [@eglen_toward_2017] and Psychology [@munafo_manifesto_2017]. However, although calls to focus on reproducibility are now commonplace on the pages of leading scientific journals [e.g. @baker_1500_2016], only a small minority of researchers are familiar with the tools and practices that enable implementing reproducibility in their workflows. Therefore, although there now is a broad consensus that efforts to improve reproducibility are important, materials instructing researchers in using them are lacking.

@@

Consider for a while if you have ever struggled with the following questions [@ihle_striving_2017, p.2]: Have you ever found a mistake in your results without knowing what caused it? Forgot what analyses you have already done and how (and possibly, why)? Lost datasets or information about the cases or variables in a dataset? Struggled in redoing statistical or computational analyses when new data became available? Had difficulty understanding what data to use or how in a project that you inherited from another researcher? Answering any of these questions in the affirmative suggests that your work might benefit from improving reproducibility [@ihle_striving_2017].

But what exactly is reproducibility? Reproducibility can be defined as follows: "A research project is computationally reproducible if a second investigator (including you in the future) can recreate the final reported results of the project, including key quantitative findings, tables, and figures, given only a set of files and written instructions." [@kitzes_practice_2017]. In the context of experimental Psychology, for example, a "project" could be an experiment or a set of experiments investigating a theory or hypothesis, "reported results" could be figures or statistical analyses reported in a conference presentation or a manuscript.

This definition makes an important distinction between *reproducibility* and *replicability*. Methods and results sections in traditional journal articles have focused on ensuring replicability by giving detailed instructions on how to repeat the experiment and collect a data set similar to the original one. However, little or no emphasis is placed on issues related to reproducibility, such as details of the data processing pipeline and statistical model exploration, although these issues are often more important to the reliability of the scientific results than are their end results (e.g. reported *p*-values; @leek_statistics:_2015). To put it more generally, replicability is a broad conceptual issue concerning the epistemology of scientific claims. Reproducibility, on the other hand, is a technical requirement for allowing others (including researchers themselves) to assess the reliability of a specific set of empirical or computational results.

Further, although they are often erroneously conlated [@open_science_collaboration_estimating_2015; @peng_reproducible_2011] reproducibility and replicability are independent of one another. Although the distinction between the two may be somewhat blurred in purely computational areas (e.g. simulation studies; @peng_reproducible_2011), it is clear that an experiment can be replicated even if the materials of the original study are not available for replicability. Likewise, the materials of an experiment might allow its results to be reproduced, yet the underlying scientific hypothesis might not be replicable. Therefore, it is important to keep these two concepts distinct; in the remainder of the manuscript we explicitly only discuss reproducibility.

Given this broad definition of reproducibility, it is easy to recognize that many projects fall between the extremes of completely replicable and not replicable at all. A completely reproducible project would provide consumers the entire raw data set, and detailed instructions to analyze it (usually as unambigous computer code). This degree of reproducibility is very rare in Psychology [@wicherts_poor_2006; @vanpaemel_are_2015], although data sharing has become more common in recent years in journals that award authors who do so with special badges [@kidwell_badges_2016]. However, even when authors (report to) share data, the shared data sets sometimes turn out to be incorrect, not usable, incomplete, or not available at all [@kidwell_badges_2016]. This suggests that reproducibility requires more than good intentions. What, then, are some specific challenges to reproducibility?

@@

Evidence suggests that the level of reproducibility within Psychology is not very high. One survey of leading Psychology journals found that 34-58% of articles published between 1985 and 2013 had at least one inconsistently reported *p*-value [@nuijten_prevalence_2015]: Upon recalculation, the authors couldn't obtain the same *p*-values from the reported test statistics, and therefore these 34-58% of articles were, to some degree, non-reproducible (the inferential statistics could not be reproduced).

Why, do so many research products fall short of the basic scientific standard of reproducibility? We suggest one reason is the poor level at which researchers organize, curate, and collaborate on the research assets---e.g. data and code used to analyze it---which support the product's conclusions. In one study, researchers asked authors of 141 Psychology articles (with a total of 249 studies) to share their data for reanalysis. 73% of the authors refused to share their data, leading the study's authors to suggest that the considerable efforts involved in cleaning datasets at the final stages of publication may make data sharing unattractive to many [@wicherts_poor_2006; see also @vanpaemel_are_2015]. Vanpaemel et al. highlight some reasons provided by authors for not sharing their data:

>"To our surprise, some authors are apparently willing to share, but have no easy access to their own data or have lost their data altogether, due to computer crashes or collaborators having left the university. Many authors cite a lack of time as a reason not to share, and note that sharing their data would take too much effort, which is probably due to poor documenta- tion and storage practices." (p. 2-3)

In short, scientists usually excel at *producing* research assets, but commonly fail at *curating* them (a topic commonly known as Digital Asset Management; @jacobsen_implementing_2012). This state of affairs is not surprising because researchers are trained to do research, but receive little or no training in formal methods or accepted gold standards for curating their research assets. For simple[^simple] projects, such as writing a solo-author manuscript, this has not been a problem. However, with increasing complexity of research designs, number of collaborators, research sites, and technical skills required in modern scientific workflows, not knowing how to keep track of materials can lead to serious problems, as detailed above.

Fortunately for the empirical sciences, challenges related to organizing and curating materials ("digital assets") across time, space, and personnel have been solved to a high standard in computer science with Version Control Systems (VCS). In the remainder of this article, we introduce a popular VCS called Git, and illustrate its use in the scientific workflow with a hypothetical example project. In the tutorial below, we show how to use the Git VCS by using text commands from the computer's command line. After introducing the fundamentals of Git's VCS functions from the command line, we also show how to use Git easily with a Graphical User Interface (GUI) implemented in the popular R Studio Integrated Development Environment (IDE).

@@

It is important to understand that unlike an operating system's (OS) default file viewer, such as Finder (Apple computers) or File Explorer (?? Windows), Git[^git-name] is not a standalone program for navigating files and folders on a computer. Instead, it adds functionality to the OSs existing file system, by making available a specific set of functions--either executed from the command line, or through a graphical user interface (GUI) / integrated development environment (IDE)--that allow taking snapshots of the project and its files, and distributing the work across multiple computers and users. To begin using Git, users must first download and install the Git software on their computers. Git is free and open source, works on Windows, Macintosh, and Linux OSs (among others), and is used by major software developers such as Microsoft, Google, and Facebook. It can be freely downloaded at <https://git-scm.com/>. Before we explain the basics of using Git, we present brief installation instructions for Windows and Apple users.

@@

[^git-version]: As of the writing of this article, the current version of Git is 2.13.1.
[^git-name]: The creator of Git, Linus Torvalds, named Git after himself as "the stupid content tracker" (@mcmillan_after_2005; <https://git-scm.com/docs/git.html>)
[^simple]: By "simple", we only mean that the technical aspects of the project, related to curating materials related to it, are simple. We of course do not suggest that there might be anything theoretically or scientifically simple in such projects.
